\documentclass[12pt]{report}
 %include polycode.fmt
\input{preamble}
\begin{document}
\input{document.preamble}

\chapter{Uncurrying}
\label{ref_chapter_uncurrying}
%% \emph{Describes our optimization for collapsing intermediate
%% closures. Our choice of representation is analyzed to
%% show how it facilitates this optimization. We should show one
%% closure can be eliminated from a program and how the optimization
%% is applied over and over until a fixed point is reached. The format
%% for this section will vary from the other two.}

Functional languages permit definitions in two styles: \emph{curried}
and \emph{uncurried}. A curried function can be \emph{partially
  applied} --- it does not need to be given all of its arguments at
once. A function that takes the remaining arguments results from such
an application. An \emph{uncurried} function, however, must be given
all of its arguments at once. It cannot be partially applied. For
example, these Haskell fragments define @adder@ in curried style and
@divider@ in uncurried style:

\begin{code}
adder :: Int -> Int -> Int
adder a b = a + b

divider :: (Float, Float) -> Float
divider (a,b) = a / b
\end{code}

\noindent
The first definition lets us easily define specialized versions of @adder@, 
such as @add1@:

\begin{code}
add1 :: Int -> Int
add1 = adder 1
\end{code}

When applied to a single argument, @adder@ returns a function that we can re-use over and over. We cannot as easily define a simliar specialized function with @divider@.

%% Why is this a problem? Need more motivatin
The implementation of partial application, however, does come at a
cost. Each partial application requires that we construct a closure
over the arguments captured so far. That closure represents a function specialized
to the arguments given so far. In general, we don't know the address of the function
it contains when compiling -- only at run-time. Therefore, when the closure is
applied to the rest of the arguments, we cannot generate code that jumps to a
known address. Instead, we must look at the address in the closure at run-time and 
then jump. 

Because each function application @f x@ may result in another
function, the most general implementation strategy makes \emph{every}
application result in a closure. The compiler need only generate code
that inspects the closure constructed and jumps to the address
indicated. When a curried function is applied to all of its arguments
at once (e.g., @adder 1 2@), we get a chain of function calls where
most construct a closure and immediately return. It would be more
efficient to collect all arguments at once and immediately jump to the
function body. \emph{Uncurrying} is the transformation we use to turn 
fully-applied curried functions into direct calls to the function body.

%% TODO: Talk about how we can look for fully-applied forms
%% as a special case, but that is sub-optimal

%% TODO: What is an example of a fully-applied function that we cannot
%% recognize syntatically (very easily)?

\section{Example of Desired Optimization}

Recall from Section \ref{ref_foo} our definition of @foldr@:

\begin{code}
foldr :: (a -> b -> b) -> b -> [a] -> b
foldr f b (a:as) = foldr f (f a b) as
foldr f b []     = b
\end{code}

which compiles to the following blocks in our MIL:

\begin{code}

\end{code}

\section{Implementation}
\section{Reflection}
\subsection{Prior Work}

\end{document}
